---
title: "From tidy climate inputs to annual CWD and surplus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From tidy climate inputs to annual CWD and surplus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

project_root <- if (file.exists("R/simulate_snow.R")) {
  "."
} else {
  ".."
}

library(dplyr)
library(ggplot2)
library(lubridate)
library(purrr)
library(tibble)
library(tidyr)

source(file.path(project_root, "R", "convert_et.R"))
source(file.path(project_root, "R", "simulate_snow.R"))
source(file.path(project_root, "R", "calculate_annual_cwd.R"))
```

This vignette demonstrates the retained core workflow without requiring any
external climate files. It starts with synthetic nested tidy tables that mimic
one longitude-slice file written by the map2tidy adapter, then performs the
same major transformations used for MSWEP precipitation, ERA5-Land
temperature, and ALEXI evapotranspiration:

1. join the nested daily climate inputs by grid cell and date;
2. convert temperature and ET to the units used by the water-balance step;
3. partition total precipitation into rain and snow and simulate snowmelt;
4. calculate daily liquid-water input minus ET;
5. calculate annual maximum CWD and match it with the latest preceding
   cumulative-surplus maximum.

No site data, soil properties, rooting-depth conversion, threshold analysis,
or extreme-value distribution is involved.

## Synthetic map2tidy-like inputs

A real tidy checkpoint has one outer row per grid cell and a nested data
column containing the daily time series. The three synthetic inputs below use
the source variable names expected for MSWEP (precipitation), ERA5-Land
(t2m, in kelvin), and ALEXI (et, in MJ m^-2^ d^-1^).

```{r synthetic-inputs}
set.seed(42)

dates <- seq(as.Date("2003-01-01"), as.Date("2008-12-31"), by = "day")
locations <- tibble(
  lon = rep(7.05, 4),
  lat = c(45.85, 45.95, 46.05, 46.15),
  elv = c(250, 400, 550, 700),
  cell = seq_len(4)
)

make_synthetic_cell <- function(lat, cell) {
  n <- length(dates)
  seasonal <- sin(2 * pi * (yday(dates) - 105) / 365.25)
  temp_c <- 7.5 + 12.5 * seasonal - 0.35 * (lat - 46)
  wet_probability <- pmax(
    0.08,
    0.30 - 0.16 * pmax(seasonal, 0) + 0.015 * (cell - 2)
  )
  precipitation <- ifelse(
    runif(n) < wet_probability,
    rgamma(n, shape = 1.6, scale = 4.2),
    0
  )
  et_energy_mj <- pmax(
    0.8,
    4.2 + 4.5 * pmax(seasonal, -0.5) + 0.15 * cell
  )

  tibble(
    time = dates,
    precipitation = precipitation,
    t2m = temp_c + 273.15,
    et = et_energy_mj
  )
}

synthetic_cells <- map2(locations$lat, locations$cell, make_synthetic_cell)

precipitation_tidy <- locations |>
  transmute(
    lon,
    lat,
    data = map(synthetic_cells, ~select(.x, time, precipitation))
  )
temperature_tidy <- locations |>
  transmute(
    lon,
    lat,
    data = map(synthetic_cells, ~select(.x, time, t2m))
  )
et_tidy <- locations |>
  transmute(
    lon,
    lat,
    elv,
    data = map(synthetic_cells, ~select(.x, time, et))
  )

precipitation_tidy
precipitation_tidy$data[[1]] |> slice_head(n = 4)
```

## Join and transform the daily drivers

The synthetic ERA5-Land values are converted from kelvin to degrees Celsius.
The ALEXI-like daily energy is converted from MJ m^-2^ d^-1^ to J m^-2^ d^-1^
and then to ET in mm d^-1^. MSWEP total precipitation is partitioned by the
temperature threshold inside simulate_snow().

```{r water-balance}
drivers <- temperature_tidy |>
  rename(data_temperature = data) |>
  left_join(
    precipitation_tidy |> rename(data_precipitation = data),
    by = c("lon", "lat")
  ) |>
  left_join(
    et_tidy |> rename(data_et = data),
    by = c("lon", "lat")
  ) |>
  mutate(
    data = pmap(
      list(data_temperature, data_precipitation, data_et, elv),
      function(temperature, precipitation, et, elevation) {
        temperature |>
          left_join(precipitation, by = "time") |>
          left_join(et, by = "time") |>
          mutate(
            temp = t2m - 273.15,
            et_mm = convert_et(
              et * 1e6,
              temp,
              elevation
            ),
            prec = precipitation
          ) |>
          select(time, temp, prec, et_mm) |>
          simulate_snow()
      }
    ),
    balance = map(
      data,
      ~transmute(
        .x,
        time,
        bal = liquid_to_soil - et_mm
      )
    )
  ) |>
  select(lon, lat, data, balance)

drivers$balance[[1]] |> slice_head(n = 6)
```

The daily balance is positive when rain plus snowmelt exceeds ET and negative
when atmospheric water loss exceeds liquid input to the soil.

```{r daily-balance-plot}
daily_example <- drivers$data[[1]] |>
  select(time, liquid_to_soil, et_mm) |>
  left_join(drivers$balance[[1]], by = "time")

ggplot(daily_example, aes(time, bal)) +
  geom_hline(yintercept = 0, color = "grey60") +
  geom_line(linewidth = 0.25) +
  labs(
    x = NULL,
    y = "Daily water balance (mm d-1)",
    title = "Synthetic liquid input minus evapotranspiration"
  ) +
  theme_minimal()
```

## Annual maximum CWD and preceding surplus

calculate_annual_cwd() adds the first-year spin-up, delegates the daily
deficit and surplus calculations to cwd::cwd(do_surplus = TRUE), removes
padding, and returns one annual table per grid cell.

```{r annual-cwd}
annual_output <- drivers |>
  mutate(
    cwd_result = map(
      balance,
      ~calculate_annual_cwd(
        .x,
        years = 2003:2008,
        return_cumulative_surplus = FALSE
      )
    ),
    annual_cwd = map(cwd_result, "annual_cwd"),
    cwd_status = map_chr(cwd_result, "status")
  ) |>
  select(lon, lat, cwd_status, annual_cwd)

annual_values <- annual_output |>
  unnest(annual_cwd)

annual_values
```

The two annual quantities can be viewed as time series for each grid cell.

```{r annual-time-series}
annual_long <- annual_values |>
  select(lat, year, cwd_mm, preceding_surplus_mm) |>
  pivot_longer(
    c(cwd_mm, preceding_surplus_mm),
    names_to = "quantity",
    values_to = "mm"
  )

ggplot(
  annual_long,
  aes(year, mm, color = factor(lat), group = interaction(lat, quantity))
) +
  geom_line() +
  geom_point(size = 1.2) +
  facet_wrap(~quantity, scales = "free_y", ncol = 1) +
  labs(
    x = NULL,
    y = "Water equivalent (mm)",
    color = "Latitude"
  ) +
  theme_minimal()
```

The final results step retains only their relationship.

```{r relationship}
relationship <- annual_values |>
  filter(
    is.finite(cwd_mm),
    is.finite(preceding_surplus_mm),
    cwd_mm > 0,
    preceding_surplus_mm > 0
  )

ggplot(
  relationship,
  aes(preceding_surplus_mm, cwd_mm, color = factor(lat))
) +
  geom_point(size = 2) +
  geom_smooth(
    aes(group = 1),
    method = "lm",
    formula = y ~ x,
    color = "black",
    se = FALSE
  ) +
  labs(
    x = "Preceding maximum cumulative surplus (mm)",
    y = "Annual maximum CWD (mm)",
    color = "Latitude"
  ) +
  theme_minimal()
```
